---
title: "Method of Moments & Least Squares Estimation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	fig.align = "center"
)
```

## Notes from Handout (MoM)


## Least Squares Estimation

$Y_t - \mu = \phi(Y_t-\mu) + e_t$

Differences: $Y_t - \mu = \phi(Y_t-\mu)$

This is a function of $\phi$ and $\mu$ (parameters you're trying to estimate).

$S(\phi, \mu) = \sum_{t-2}^n [(Y_t - \mu) - \phi(Y_{t-1} - \mu)]^2$

To minimize sum of squares, take partial derivatives with respect to each term.

Equations 7.2.. in the book for how to do the differentiation and derive the general formula


### MA(1)

$Y_t = e_t - \theta e_{t-1}$

$Y_t = e_t - \theta Y_{t-1} - \theta^2 Y_{t-2} - \theta^3 Y_{t-3} - ...$

Minimize the series $\theta Y_{t-1} - \theta^2 Y_{t-2} - \theta^3 Y_{t-3} - ...$


**Homework Question 1**

Colton's question about characteristic equations
$\phi (x)  = 1 - \phi \beta x$

$AB \rightarrow (1 - \phi_1 \beta)Y_t$

$Diff \rightarrow Y_t - T_{t-1}$ or written as $[1 - \beta]Y_t$

Second difference

$(1-\beta)(Y_t - Y_{t-1})$

$Y_t - T_{t-1} - Y_{t-1} + Y_{t-2}$

$Y_t - 2Y_{t-1} + Y_{t-2}$

$[1 - 2\beta + \beta^2]$

$(1 - \beta)^2Y_t$

**Example ARIMA(1,1,1)**

$[1 - 0.6\beta](1-\beta)Y_t = (1 - 0.4\beta)e_t$

$(1 - 0.6\beta)[Y_t - Y_{t-1}] = e_t - 0.4e_{t-1}$

$Y_t - 0.6Y_{t-1} - Y_{t-1} + 0.6Y_{t-2} = e_t - 0.4e_{t-1}$

$Y_t - 1.6Y_{t-1} + 0.6Y_{t-2} = e_t - 0.4e_{t-1}$

$Y_t = 1.6Y_{t-1} - 0.6Y_{t-2} + e_t - 0.4e_{t-1}$


**Another Example** Similar to Problem B

Identify p,d & q for $Y_t = 1.6Y_{t-1} - 0.6Y_{t-2} + e_t - 0.4e_{t-1}$

$AR(2) \rightarrow 1.6, -0.6$  so to determine stationarity, $1.6 + (-0.6) = 1$.  Since this is greater than 1, this process is not stationary.

$Y_t -1.6_{t-1} + 0.6Y_{t-2} = e_t - 0.4e_{t-1}$

$(1 - 1.6\beta + 0.6\beta^2)Y_t$

$(1 - 0.6\beta)(1 - \beta)Y_t = (1 - 0.4\beta)e_t$

So take this equation and let $W_t = Y_t - Y_{t-1}$ and then write the equation as 

$(1 - 0.6\beta)W_t = (1 - 0.4\beta)e_t$

$W_t - 0.6W_{t-1} = (1 - 0.4\beta)e_t$

$W_t = 0.6W_{t-1} + e_t - 0.4e_{t-1}$

$Y_t - Y_{t-1} = 0.6Y_{t-1} - 0.6Y_{t-2} + e_t - 0.4e_{t-1}$

$Y_t = (1 + 0.6)Y_{t-1} - 0.6Y_{t-2} + e_t - 0.4e_{t-1}$

This final equation equals $Y_t = (1 + \phi)Y_{t-1} - \phi Y_{t-2} + e_t - 0.4e_{t-1}$


## Maximum Likelihood Estimation (MLE)

  * We will assume our parameters are independent and normally distributed.
  * Think Bayesian statistics, the values are known and we are estimating the parameter



**Normal Density Funciton**

$f(x) = \frac{1}{\sqrt{2\pi \sigma^2}} e^{\frac{(x-\mu)^2}{2\sigma^2}}$

### Class Examples

 * color dataset from TSA
 
```{r class_ex}

data(color)
autoplot(color)


```

```{r class_ex2}

arima(color)

```

ARIMA (2,0,0)

```{r class_ex3}

arima(color, order = c(2,0,0), method = "ML")

```

**Another Example**

```{r ex2}

data(ar2.s)
autoplot(ar2.s)

```

```{r ex2_acf}
plot(ggAcf(ar2.s))


```

```{r ex2_pacf}
plot(ggPacf(ar2.s))

```

```{r ex2_eacf}
eacf(ar2.s)

```

```{r ex2_arima}
arima(ar2.s, order = c(2, 0, 0))

```

**Oil Price Data Example**

```{r ex3}
data("oil.price")
autoplot(oil.price)

```

```{r ex3_log}
autoplot(log(oil.price))

```

```{r ex3_acf}
plot(ggAcf(oil.price))

```

```{r ex3_pacf}
plot(ggPacf(oil.price))

```

```{r ex3_eacf}
eacf(oil.price)

```

Original plot of log(oil.price) doesn't look stationary, look at differencing to achieve stationarity

```{r ex3_difference}
diff = log(oil.price)[2:241] - log(oil.price)[1:240]
plot(ggAcf(diff))
```

```{r ex3_diff_pacf}
plot(ggPacf(diff))

```

## Modeling Process

What steps do you take to model a given time series?

  * Plot the time series
  * Look for seasonality, stationarity
  * Constant Variability (Transformation if non-constant variance)
  * Look at ACF, PACF and EACF to find p,d,q
  * Fit the model
  * Evaluate the estimates and are they significant
  * Are the residuals normally distributed and independent
  * And don't forget about the runs test, shapiro-wilkes test for normality, normal probability plots....
  
  
  
### Color Data Example

```{r oil_example}
library(TSA)
data("oil.price")
plot(oil.price)
plot(log(oil.price))
plot(diff(log(oil.price)))

```


```{r oil_diagnostics}

acf(diff(log(oil.price)))
pacf(diff(log(oil.price)))
eacf(diff(log(oil.price)))


```

```{r oil_fit}

ma1_oil = arima(log(oil.price), order = c(0, 1, 1))
ma1_oil

acf(ma1_oil$residuals)
pacf(ma1_oil$residuals)
eacf(ma1_oil$residuals)

```

```{r oil_testing}
#Possible outliers causing the shapiro test to fail 
shapiro.test(ma1_oil$residuals)
hist(ma1_oil$residuals)

#runs test
runs(ma1_oil$residuals)

# Ljung-Box Test
tsdiag(ma1_oil)

```

