---
title: "Trends"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

## Introduction

Model $Y_t = \mu_t + X_t$

  * $\mu_t$ = trend
  * $X_t$ = random error
  
If $E(X_t) = 0$ (zero mean process), then $E(Y_t) = E(\mu_t + X_t) = \mu_t$

## Common Models

**Linear**

Increasing/Decreasing at a constant rate.

\[
\mu_t = \beta_0 + \beta_1t
\]

**Quadratic**
\[
\mu_t = \beta_0 + \beta_1t + \beta_2t^2$
\]

......and so on for higher order models

**Cycles**

Use sin and cos functions to model trends with periodicity

$\mu_t = \beta_0 + \sum(\alpha; cos\omega; t + \beta_jsin\omega_j; t)$


Another way to remove a trend is a method called differencing.  This is simply taking differences such as:

$Y_t - Y_{t-1}$


Consider $Y_t = \mu + X_t$.  The trend is the constant $\mu$.  Often $\bar{Y} = \frac{1}{n}\sum_{i=1}^n Y_i$ is an estimator for $\mu$. 

*Formula 2.2.7 in the book*

$Var(\bar{Y}) = Var(\frac{1}{n}y_1 + \frac{1}{n}y_2 + ... + \frac{1}{n}y_n)$
$Var(\bar{Y}) = \frac{1}{n^2}\sum_{t=1}^n Var(Y_i) + 2\sum_{t=2}^n\sum_{j=1}^{t-1} c_tc_j\gamma_{t,j}$

$= \frac{1}{n^2}[n\gamma_o] + \frac{2}{n^2}[\gamma_{2,1} + \gamma_{3,1} + \gamma_{3,2} + \gamma_{4,1} + \gamma_{4,2} + \gamma_{4,3} +... + \gamma_{n,n-1}]$

*Note:* $\gamma_o = Var(Y_t)$
      $\gamma_{t,t}= \gamma_{j,j} = \gamma_{o}$
      

## In Class Code Excercises

```{r exercise_1}
library(tidyverse)
library(TSA)
data(rwalk)
plot(rwalk, type = "o")

model1 = lm(rwalk~time(rwalk))
summary(model1)

plot(y=rstudent(model1), x=as.vector(time(rwalk)), type="o")
acf(rstudent(model1))

```



```{r excercise_2}

# Dubuque, Iowa Temperature Data
data(tempdub)
modeldub=lm(tempdub~time(tempdub))
summary(modeldub)
plot(y=rstudent(modeldub), x=as.vector(time((tempdub))), type="o")
acf(rstudent(modeldub))
acf(tempdub) # notice the pattern 12 time periods apart

# Seasonal Means Model
month.=season(tempdub) # period added to help table display
model2 = lm(tempdub~month.-1) # The -1 removes the intercept
summary(model2)
acf(rstudent(model2))

# Alternative Fit
model3=lm(tempdub~month.) # January is automatically dropped
summary(model3)
acf(rstudent(model3))

```


For a series $Y_1, Y_2, Y_3, ..., Y_n$:

> $Y_t = \mu_t + X_t$

$\mu_t \rightarrow$ find model that will ignore this term

$X_t \rightarrow$ this is the variable we would like to model

## What is Independence?

Rolling dice example in class

### Stationary Process

Expected Value of Rolling Dice

\[
\begin{eqnarray}
E(Y_t) & = & E(r_1 + r_2 + ... + r_t)\\
& = & 4.5t\\
u_t & = & Y_1 - Y_2\\
& = & r_t
\end{eqnarray}
\]

Look at differences instead of the expected value

```{r dice_rolling_example}

Roll = c(3, 4, 5, 6, 3, 1, 7, 8, 6, 2, 5, 2, 7, 1, 5, 3, 2, 5, 3, 2, 4, 1, 4, 2, 1)
timeIndex = c(1:25)

diceData = tibble(timeIndex, Roll)

dicePlot = ggplot(diceData, aes(x = timeIndex, y = Roll)) + 
  geom_point() + 
  geom_line() + 
  geom_hline(yintercept =  4.5, linetype = "dashed", color = "blue") +
  ggtitle("8-Sided Die Rolls")

plot(dicePlot)

```

```{r runs_test}
library(TSA)
library(kableExtra)

# List of values after rolling dice 25 times
Roll = c(3, 4, 5, 6, 3, 1, 7, 8, 6, 2, 5, 2, 7, 1, 5, 3, 2, 5, 3, 2, 4, 1, 4, 2, 1)

# Force to a data frame to allow as input to kable
runsTest = as.data.frame(runs(Roll, k = 4.5))

# Create a table of the output
kable(runsTest) %>% 
  kable_styling(full_width = F) %>% 
  add_header_above(header = c("Runs Test" = 6))
  


```

Consider the random walk from TSA package (Expected Value is 0)

```{r runsTest_randomWalk}
library(TSA)

data(rwalk)
win.graph(12,7,14)
plot.ts(rwalk, type="o")
abline(h=0, col="blue")

runs(rwalk)

```

## Periodic Functions

$\mu_t \rightarrow$ periodic function

$Y_t = \mu_t + X_t = \beta cos(2\pi ft + \Theta$

$\beta \rightarrow$ amplitude *($\mu_t$ oscillates between $-\beta$ and $\beta$)*

$f \rightarrow$ frequency *($\frac{1}{f}$ is the period)*

$\Theta \rightarrow$ phase shift

### Trigonometry Review

$cos(a + b) = cos(a) + cos(b) - sin(a) + sin(b)$

$\beta cos(2\pi ft + \Theta = \beta[cos(2\pi ft)cos(\Theta) - \beta\sin(2\pi ft)\sin(\Theta)]$

The only random variable in the above equation is $t$. Combining $\beta$ and $cos(\Theta)$:

$\beta_1cos(2\pi ft) - \beta_2sin(2\pi ft)$

$\beta_1 = \beta cos(\Theta)$

$\beta_1 = \frac{\beta}{cos(\Theta)}$


$\beta_2 = -\beta sin(\Theta)$

$\beta_2 = \frac{-\beta}{sin(\Theta)}$

Given these two betas:

$\frac{\beta_1}{cos(\Theta)} = \frac{-\beta}{sin(\Theta)}$

$\frac{sin(\Theta)}{cos(\Theta)} = \frac{-\beta_2}{\beta_1}$

$tan(\Theta) =  \frac{-\beta_2}{\beta_1}$

$\Theta = tan^{-1}(\frac{-\beta_2}{\beta_1})$

### Example

Cosine Example from Canvas

```{r cos_example}
library(TSA)
data("tempdub")

har.=harmonic(tempdub,1)
model4tempcos=lm(tempdub~har.)
summary(model4tempcos)

```

Raw data plotted

```{r cos_example_contd}
plot(tempdub, type="o")

#plot(ts(fitted(model4tempcos),freq=12,start=c(1964,1)),ylab='Temperatur',type='l',ylim=range(c(fitted(model4tempcos),tempdub)));
#points(tempdub)

```

Data fitted to the model.  Notice the data points at the bottom of the chart (low temperatures) are not accounted for by the model.

```{r cos_example_continued}
plot(ts(fitted(model4tempcos),freq=12,start=c(1964,1)),ylab='Temperatur',type='l',ylim=range(c(fitted(model4tempcos),tempdub)));
points(tempdub)

```

## Moving Average Process

**MA(2) Process**

$Y_t = e_t - \Theta_1e_{t-1} - \Theta_2e_{t-2}$

e's are a white noise process

> E($e_i$) = 0  *zero mean*
>
> Var($e_i$) = $\sigma_{\epsilon}^2$


$E(Y_t) = E[e_t - \Theta_1e_{t-1} - \Theta_2e_{t-2$

$= 0 + 0 + 0$

$\gamma_o = var(Y_t) = \sigma_{\epsilon}^2[1 + \Theta_1^2 + \Theta_2^2]$

**Look in notes**

```{r}
library(TSA)

win.graph(width = 5, height = 3, pointsize = 14)
data("ma2.s")
plot(ma2.s, ylab = expression("[Yt]"), type = "o")

```


